---
title: "dml"
author: "Shun"
date: "2026-06-05"
output:
  pdf_document: default
  html_document: default
---

```{r}

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE
)
```

# Run DML Analysis

```{r}
source("dml_analysis (1).R")
```

# Main Results

```{r}
knitr::kable(
  best[,c(
    "group",
    "treatment",
    "learner",
    "ATE",
    "SE",
    "p_value",
    "sig"
  )],
  caption = "Best DML Results by Treatment"
)
```

# Full Results

```{r}
knitr::kable(
  results_df[,c(
    "group",
    "treatment",
    "learner",
    "ATE",
    "SE",
    "p_value"
  )],
  caption = "All DML Estimates"
)
```

# Figure 1: ATE Estimates

```{r}
library(ggplot2)

ggplot(
  best,
  aes(
    x = reorder(treatment, ATE),
    y = ATE,
    fill = group
  )
) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Estimated Treatment Effects",
    x = "",
    y = "ATE"
  ) +
  theme_minimal()
```

# Figure 2: P-values

```{r}
ggplot(
  best,
  aes(
    x = reorder(treatment, p_value),
    y = p_value,
    fill = group
  )
) +
  geom_col() +
  geom_hline(
    yintercept = 0.05,
    linetype = "dashed"
  ) +
  coord_flip() +
  labs(
    title = "Statistical Significance",
    x = "",
    y = "P-value"
  ) +
  theme_minimal()
```

# Figure 3: Model Comparison

```{r}
ggplot(
  results_df,
  aes(
    learner,
    mse_outcome,
    fill = learner
  )
) +
  geom_boxplot() +
  labs(
    title = "Outcome Nuisance MSE by Learner",
    x = "",
    y = "MSE"
  ) +
  theme_minimal()
```

# Interpretation

The table above reports Double Machine Learning estimates of the relationship between employment outcomes and three groups of predictors:

1. Health and Social Factors
2. Financial Responsibility
3. Lagged Wellbeing

For each treatment variable, the learner with the lowest nuisance-model MSE was selected as the preferred specification.

```

