---
title: "ECON484 project"
author: "Natalie Li"
date: "2026-06-04"
output: pdf_document
---
# Predicting Employment Outcomes among Young Adults: Education, Well-being, and Machine Learning Evidence from PSID-TAS 2023.

```{r}
library(readxl)
library(dplyr)
library(ggplot2)
library(glmnet)
library(randomForest)
library(rpart)
library(rpart.plot)
library(gbm)
library(nnet)
```

```{r}
data <- read_excel("/Users/natalieli/Desktop/ECON484 Project/2023/73datas.xlsx")

dim(data)
names(data)
```

```{r}
project_data <- data %>%
  mutate(
    employed = ifelse(
      TA230239 %in% c(1,2),
      1,
      0
    )
  )
table(project_data$employed)
```

## Wellbeing
```{r}
#TA230061 happy
#TA230062 interest in life
#TA230063 satisfied
#TA230065 community
#TA230066 society better
#TA230067 people good
#TA230069 manage responsibilities
#TA230070 trusting relationships
#TA230072 confidence
#TA230074 life purpose

wellbeing_vars <- c(
  "TA230061",
  "TA230062",
  "TA230063",
  "TA230065",
  "TA230066",
  "TA230067",
  "TA230069",
  "TA230070",
  "TA230072",
  "TA230074"
)

project_data <- project_data %>%
  mutate(
    across(
      all_of(wellbeing_vars),
      ~ifelse(. == 9, NA, .)
    )
  )

project_data <- project_data %>%
  rowwise() %>%
  mutate(
    wellbeing_score =
      mean(
        c_across(all_of(wellbeing_vars)),
        na.rm = TRUE
      )
  ) %>%
  ungroup()
summary(project_data$wellbeing_score)

ggplot(
  project_data,
  aes(x = wellbeing_score)
) +
  geom_histogram(
    bins = 20
  ) +
  labs(
    title = "Distribution of Well-being Scores"
  )

project_data %>%
  group_by(employed) %>%
  summarise(
    avg_wellbeing =
      mean(wellbeing_score,
           na.rm = TRUE)
  )

#Employment vs Well-being
ggplot(
  project_data,
  aes(
    x = factor(employed),
    y = wellbeing_score
  )
) +
  geom_boxplot() +
  labs(
    x = "Employment Status",
    y = "Well-being Score"
  )
```
```{r}
#Logistic Regression
#Does well-being predict employment?
wellbeing_logit <- glm(
  employed ~ wellbeing_score,
  data = project_data,
  family = binomial
)

summary(wellbeing_logit)
```
Higher well-being is associated with a slightly higher probability of being employed, but this relationship is weak and not statistically significant in the simple logit model.

```{r}
#Tree
wellbeing_tree_data <- project_data %>%
  select(
    employed,
    all_of(wellbeing_vars)
  ) %>%
  na.omit()

colnames(wellbeing_tree_data) <- c(
  "employed",
  "Happy",
  "InterestInLife",
  "Satisfied",
  "Community",
  "SocietyBetter",
  "PeopleGood",
  "ManageResponsibilities",
  "TrustingRelationships",
  "Confidence",
  "LifePurpose"
)

wellbeing_tree <- rpart(
  employed ~ .,
  data = wellbeing_tree_data,
  method = "class"
)

rpart.plot(wellbeing_tree)

printcp(wellbeing_tree)
plotcp(wellbeing_tree)
```
The CART model failed to identify meaningful splits, suggesting that individual well-being indicators alone do not strongly distinguish employed from non-employed young adults.

```{r}
#Random Forest
set.seed(123)

wellbeing_rf <- randomForest(
  as.factor(employed) ~ .,
  data = wellbeing_tree_data,
  ntree = 500,
  importance = TRUE
)

wellbeing_rf

varImpPlot(wellbeing_rf)
```
Some well-being indicators appear moderately important, but predictive power remains weak.

```{r}
#LASSO
x_wellbeing <- model.matrix(
  employed ~ .,
  wellbeing_tree_data
)[, -1]

y_wellbeing <- wellbeing_tree_data$employed

set.seed(123)

wellbeing_lasso <- cv.glmnet(
  x_wellbeing,
  y_wellbeing,
  family = "binomial",
  alpha = 1
)

plot(wellbeing_lasso)

coef(wellbeing_lasso, s = "lambda.min")
coef(wellbeing_lasso, s = "lambda.1se")
```
lambda.1se selected no predictors



## Education
```{r}
education_vars <- c(
  "TA232393",
  "TA230990",
  "TA231000",
  "TA231001",
  "TA230986",
  "TA230988",
  "TA230685",
  "TA232394",
  "TA232396"
)

education_labels <- c(
  "HighestEducation",
  "GradeCompleted",
  "AttendedCollege",
  "CurrentCollege",
  "EducationAspiration",
  "EducationExpectation",
  "EducationHours",
  "MotherEducation",
  "FatherEducation"
)

education_tree_data <- project_data %>%
  select(
    employed,
    all_of(education_vars)
  ) %>%
  na.omit()

colnames(education_tree_data) <- c(
  "employed",
  education_labels
)
```

```{r}
# Logistic Regression
education_logit <- glm(
  employed ~ .,
  data = education_tree_data,
  family = binomial
)

summary(education_logit)
```
```{r}
#Tree
education_tree <- rpart(
  employed ~ .,
  data = education_tree_data,
  method = "class"
)

rpart.plot(education_tree)
```
```{r}
#Random Forest
set.seed(123)

education_rf <- randomForest(
  as.factor(employed) ~ .,
  data = education_tree_data,
  ntree = 500,
  importance = TRUE
)

varImpPlot(education_rf)
```
```{r}
x_education <- model.matrix(
  employed ~ .,
  education_tree_data
)[,-1]

y_education <- education_tree_data$employed

education_lasso <- cv.glmnet(
  x_education,
  y_education,
  family = "binomial",
  alpha = 1
)

plot(education_lasso)

coef(
  education_lasso,
  s = "lambda.min"
)

coef(
  education_lasso,
  s = "lambda.1se"
)
```

```{r}
# Final comparison table

wellbeing_AIC <- AIC(wellbeing_logit)
education_AIC <- AIC(education_logit)

wellbeing_tree_splits <- wellbeing_tree$cptable[nrow(wellbeing_tree$cptable), "nsplit"]
education_tree_splits <- education_tree$cptable[nrow(education_tree$cptable), "nsplit"]

wellbeing_lasso_1se <- coef(wellbeing_lasso, s = "lambda.1se")
education_lasso_1se <- coef(education_lasso, s = "lambda.1se")

wellbeing_lasso_vars <- sum(wellbeing_lasso_1se[-1, ] != 0)
education_lasso_vars <- sum(education_lasso_1se[-1, ] != 0)

wellbeing_rf_acc <- mean(
  predict(wellbeing_rf, wellbeing_tree_data) ==
    wellbeing_tree_data$employed
)

education_rf_acc <- mean(
  predict(education_rf, education_tree_data) ==
    education_tree_data$employed
)

comparison_table <- data.frame(
  Model_Group = c("Well-being", "Education"),
  Logit_AIC = c(wellbeing_AIC, education_AIC),
  Tree_Splits = c(wellbeing_tree_splits, education_tree_splits),
  LASSO_lambda_1se_Selected_Variables =
    c(wellbeing_lasso_vars, education_lasso_vars),
  Random_Forest_Training_Accuracy =
    c(wellbeing_rf_acc, education_rf_acc)
)

comparison_table
```
```{r}
ggplot(
  comparison_table,
  aes(
    x = Model_Group,
    y = Logit_AIC
  )
) +
  geom_col() +
  labs(
    title = "Logit Model Fit Comparison",
    x = "",
    y = "AIC"
  )

ggplot(
  comparison_table,
  aes(
    x = Model_Group,
    y = LASSO_lambda_1se_Selected_Variables
  )
) +
  geom_col() +
  labs(
    title = "LASSO Variable Selection Comparison",
    x = "",
    y = "Number of Variables Selected"
  )
```
Overall, the education variables provide stronger and more stable predictive information than the well-being variables. The education logit model has a much lower AIC, the education tree identifies meaningful splits, and LASSO retains several education variables under the conservative lambda.1se rule. In contrast, the well-being tree does not split and LASSO removes all well-being variables at lambda.1se. Although the random forest can fit both sets of variables, the broader pattern suggests that education is the stronger predictor of employment outcomes.
